Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-09 manuscript/manuscript.tex
small tweaks to model feature results section

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling site-to-site variation in purifying selection.}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

@@

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}A;][]{yang1994maximum, yang2000codon}. 
Under such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, lartillot2007suppression, le2008phylogenetic, rodrigue2010mutation,si2008empirical,rodrigue2014site,bloom2017identification,hilton2017phydms,bloom2014experimentally}.
The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, which leads to $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

These two strategies to account for site-to-site variation in purifying selection are not mutually exclusive. 
\jdbcomment{I'm not sure if I love the rest of this paragraph. I feel like GY94 needs to be introduce more directly---perhaps in the paragraph beginning the ``The most common strategy...'' above. Also, it's a bit unclear what ``most appropriate'' means.}
While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.
Proteins which are under strong purifying and diversifying selection may be better modeled by a combination of the two. 
For this study, we used four models from the GY94 and ExpCM families, which together represent all possible combinations of these two strategies (\ref{fig:model_feature}C).

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models. 
The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
\skhcomment{calculated using equation. Maybe change long branch length to expected asymptotic sequence divergence.}
}
\end{figure}

@@

The specific parametrization scheme of a model will control features of its stationary state. 
Comparing GY94 to GY94+$\Gamma\omega$ will change how long it takes for the model to reach the stationary state but it does not change the expected sequence divergence at the stationary state. 
Furthermore, the GY94 and GY94+$\Gamma\omega$ stationary state is exactly the same at each site in the protein. 
Site-specific stationary states modulate both the amount of time to reach and the overall divergence level at the stationary state. 
These differences arise from whether or not the site is modeled as constrained or mutationally tolerant.  
The stationary state is important to branch length estimation because differences in expected divergence will lead to differences in branch length estimation and branch lengths cannot be estimated with any accuracy when two sequences have reached the divergence level of the stationary state.

>>>>

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling site-to-site variation in purifying selection.}
(A) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(B) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(C) Substitution models can both, one, or neither of these features and we use models from the GY94 and ExpCM families to represent all possible combinations. 
}
\end{figure}

@@

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
For example, in the M5 of \cite{yang2000codon} the implied dN/dS parameter, $\omega$, follows a gamma distribution (GY94+$\Gamma\omega$, \ref{fig:model_feature}A).
Under the GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting \skhcomment{cite rodrigue paper}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters that are fit from the phylogenetic data for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Finally, these two strategies to account for purifying selection, $\Gamma\omega$ rate variation and site-specific stationary states, are not mutually exclusive. 
Some proteins may be better modeled as a combination of the two. 
Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to examine the effects of each strategy on branch length estimation.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models calculated using \ref{eq:f}. 
The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship} (\ref{eq:w_r}).
\skhcomment{ Maybe change long branch length to expected asymptotic sequence divergence.}
}
\end{figure}

@@

\skhcomment{shorten the following sections}
The specific parametrization scheme of a model will control features of its stationary state. 
Comparing GY94 to GY94+$\Gamma\omega$ will change how long it takes for the model to reach the stationary state but it does not change the expected sequence divergence at the stationary state. 
Furthermore, the GY94 and GY94+$\Gamma\omega$ stationary state is exactly the same at each site in the protein. 
Site-specific stationary states modulate both the amount of time to reach and the overall divergence level at the stationary state. 
These differences arise from whether or not the site is modeled as constrained or mutationally tolerant.  
The stationary state is important to branch length estimation because differences in expected divergence will lead to differences in branch length estimation and branch lengths cannot be estimated with any accuracy when two sequences have reached the divergence level of the stationary state.
